The yaq Project: \\ Another Protocol for Scientific Instrumentation

### 2022-08-02 paper.tex

\documentclass{article}
\usepackage[letterpaper, margin=1in]{geometry}

\title{The yaq Project: a Protocol for Scientific Instrumentation}
\author{Kyle F. Sunden,
  John C. Wright,
  Blaise J. Thompson}

\begin{document}

@@

\section{Abstract}

@@

Instrumentation development is a key part of the scientific enterprise.
As the scientific industry has advanced, novel instruments are typically constructed of many individual components, purchased and home-built. %scientists are increasingly able to directly purchase pieces of their instruments.
Orchestration software must communicate with each hardware component in the course of a scientific experiment.
This can involve utilization of many protocols: NI DAQmx [CITE], SCPI [CITE], ModBus [CITE], PICam [CITE], Thorlabs APT [CITE], among many others.
The challenge of integrating all of these is a frustrating piece of the modern instrument development process.
Weeks can be spent just integrating one new component into an existing project.
Code reusability is typically poor, and technical debt grows quickly in academic and educational contexts.
Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.

@@

As smaller research labs have grown in experimental complexity, many hardware abstraction layers have been produced by individual labs.
Most small custom research instrumentation continues to rely on monolithic home-built orchestration software which has hardcorded protocol support for each particular connected device.
These monolithic applications tend to be very inflexible and difficult to develop.
In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity [cite many].
While this growth is encouraging, many of these are limited by their focus on particular types of hardware or particular experimental domains.
To our knowledge, none of these projects take a strict protocol-based approach.

@@

The yaq protcol is client-server network based.
To communicate with a given piece of hardware, a client connects to a particular TCP port either locally or to a remote machine over the network.
Once connected, the client will recieve an exhaustive description of what capabilities that particular hardware interface supports.
Clients can connect to many TCP ports and each TCP port can simultaniopusly communicate with many clients.
While Avro RPC itself is agnostic to transport layer, yaq specifies TCP sockets as the transport layer.
TCP socket libraries are ubiquitous, allowing for clients (or daemons) to be implemented in a variety of languages if required.
The use of TCP also works across standard networks, allowing for simple multi-machine instruments.

@@

In practice, each hardware interface is exposed to all clients as a set of well-described function calls.

\section{Traits}

Each component of a scientific instrument may have many parameters.
Similar components typically have similar parameters, but there may be incidental differences between each implementation.
For example, consider two different monochromators.
The driver for monochromator A may expose the function "set wavelength", while the driver for monochromator B exposes "set angle".
Any experimentalist wishing to switch from monochromator A to B must first go through their entire codebase to fix all of the small differences in the driver implementation.
yaq is designed using an (optional) compositional system that helps enforce consistency between different kinds of hardware.

yaq defines "traits": collections of behaviors which are logically grouped together.
Each trait specifies a set of configuration options, state entries, and methods that a daemon can implement.
Daemons, then, can be composed of multiple traits.
Clients can trust that daemons which implement a given trait will behave in similar ways.
For example, any daemon that implements the "is-position" trait must expose the methods "get\_position" and "set\_position".
It becomes easy to write generic clients that work with any daemon that has certain traits.
Of course, daemons are always allowed to implement additional configuration, state, and methods in addition to those which are implied by their traits.

\section{Daemons}

In yaq, each component of an instrument is supported by a tiny lightweight process that runs in the background of your computer: a daemon.
Since each component has its own process, each daemon can be developed separately.
Using yaq, a typical instrument might have several daemons each supporting a particular component.
Daemon A might support a motor, daemon B a sensor, and daemon C a light source.
To do an experiment, a control program (a "client") must send commands to each of these daemons.
Each of the daemons is a separate application running in its own process, and importantly the client is also a separate application.

The separation between each daemon and client makes the yaq framework less fragile than monolithic applications.
The yaq framework is language agnostic.
For example, a daemon written in Python might be controlled by a client implemented using LabVIEW.
In yaq, each component of an instrument can be developed and distributed separately.
For example, two very different instruments might happen to use the same temperature sensor.
Because the temperature sensor daemon is its own independent program, both instruments can benefit from the same daemon.
As yaq grows, the "ecosystem" of existing daemons means that future instruments become easier and easier to develop.

@@

\section{Clients}

yaq's RPC design invites the creation of multiple client libraries specialized for specific purposes.
Again, daemons can communicate simultaniously with multiple clients.
For a yaq instrument it is normal to have multiple clients running simultaniously, each focused on a particular purpose.
This modularity speeds development and increases overall flexibility of the system.
We have created a few client libraries which we will discuss here.

@@

\section{Conclusions}

\clearpage
\bibliographystyle{unsrt}
\bibliography{references}

>>>>

%  1)  latex  paper
%  2)  bibtex paper
%  3)  latex  paper
%  4)  latex  paper
%
\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1}

\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{bm}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}
\usepackage{etoolbox}

\makeatletter
\def\@email#1#2{%
 \endgroup
 \patchcmd{\titleblock@produce}
  {\frontmatter@RRAPformat}
  {\frontmatter@RRAPformat{\produce@RRAP{*#1\href{mailto:#2}{#2}}}\frontmatter@RRAPformat}
  {}{}
}%
\makeatother
\begin{document}

\preprint{AIP/123-QED}

\title{The yaq Project: \\ Another Protocol for Scientific Instrumentation}
\author{Kyle F. Sunden}
\author{John C. Wright}
\author{Blaise J. Thompson}
\affiliation{
  University of Wisconsin--Madison
}
\email{blaise.thompson@wisc.edu}

\date{\today}

\begin{abstract}
An article usually includes an abstract, a concise summary of the work
covered at length in the main body of the article. It is used for
secondary publications and for information retrieval purposes.
\end{abstract}

@@

@@

Instrumentation development is a key part of the scientific enterprise.
Novel instruments are typically constructed of many individual components, purchased and home-built.
Orchestration software must communicate with each hardware component in the course of a scientific experiment.
This can involve utilization of many protocols: NI DAQmx [CITE], SCPI [CITE], ModBus [CITE], PICam [CITE], Thorlabs APT [CITE], among many others.
The challenge of integrating all of these is a frustrating piece of the modern instrument development process.
Weeks can be spent just integrating one new component into an existing project.
Code reusability is typically poor, and technical debt grows quickly in academic and educational contexts.
Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.

@@

As smaller research labs have grown in experimental complexity, many individual labs have created domain-specific orchestration software.
Most small custom research instrumentation continues to rely on monolithic software which has hardcorded protocol support for each particular connected device.
These monolithic applications tend to be very inflexible and difficult to develop.
In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity [cite many].
While this growth is encouraging, many of these are limited by their focus on particular types of hardware or particular experimental domains.
To our knowledge, none of these projects take a strict protocol-based approach.

@@

\subsection{Architecture}

This section has to introduce the ideas of clients, daemons, and network traffic.
Should get the multilingual idea here.

Like EPICS and TANGO, the yaq protcol is client-server network based.
To communicate with a given piece of hardware, a client connects to a particular TCP port either locally or to a remote machine over the network.
Clients can connect to many TCP ports and each TCP port can simultaneously communicate with many clients.

Since each component has its own process, each daemon can be developed separately.
Daemon A might support a motor, daemon B a sensor, and daemon C a light source.
To do an experiment, a control program (a "client") must send commands to each of these daemons.
Each of the daemons is a separate application running in its own process, and importantly the client is also a separate application.

\subsection{Avro}

\begin{figure}
  \includegraphics[width=\columnwidth]{./figures/methods.png}
  \caption{  \label{fig:methods} A figure caption. The figure captions are automatically numbered.}
\end{figure}

Once connected, the client will receive an exhaustive description of what capabilities that particular hardware interface supports.
Among this description is a machine-readable summary of what the signature of each remote procedure is.
Figure \ref{fig:methods} summarizes what this might look like.
Here we see three remote procedures, each with a described set of arguments ``request'' and a described response.
As the figure shows, these signatures are typed.
This is only a small piece of the information a client receives.

While Avro RPC itself is agnostic to transport layer, yaq specifies TCP sockets as the transport layer.
TCP socket libraries are ubiquitous, allowing for clients (or daemons) to be implemented in a variety of languages if required.
The use of TCP also works across standard networks, allowing for simple multi-machine instruments.

@@

\subsection{Traits}

\begin{figure}
  \includegraphics[width=\columnwidth]{./figures/traits.png}
  \caption{  \label{fig:traits} A figure caption. The figure captions are automatically numbered.}
\end{figure}

Each component of a scientific instrument may have many parameters.
Similar components typically have similar parameters, but there may be incidental differences between each implementation.
For example, consider two different monochromators.
The driver for monochromator A may expose the function "set wavelength", while the driver for monochromator B exposes "set angle".
Any experimentalist wishing to switch from monochromator A to B must first go through their entire codebase to fix all of the small differences in the driver implementation.
yaq is designed using an (optional) compositional system that helps enforce consistency between different kinds of hardware.
We call this system ``traits''.

The most basic feature of each trait is a collection of well-defined method signatures, as illustrated by Figure \ref{fig:traits}.
Here, ``has-position'' is a trait and ``set\_position'' and ``get\_position'' are two method signatures defined by that trait.
For brevity we only show method names here, but in reality each method signature is fully described as discussed in the previous section.
Any protocol that wants to implement the ``has-position'' trait must then support those two method signatures.
Importantly, protocols can also implement arbitrary additional methods that are not defined by any trait.

The trait system was introduced in pursuit of our primary goal of easing the client development process.
Clients can trust that protocols that implement a given trait will behave in similar ways.
In Section \ref{sec:clients} we will show how this system makes it easy to write generic clients for a given experiment.
Each trait is carefully defined through prose and examples by an enhancement proposal system that helps the community to gain consensus and serves as living documentation.
These enhancement proposals can be found on the yaq website.

\subsection{Daemons}

\begin{figure}
  \label{fig:daemons}
  \includegraphics[width=\columnwidth]{./figures/daemons.png}
  \caption{ A figure caption. The figure captions are automatically numbered.}
\end{figure}

In yaq, each component of an instrument is supported by a tiny lightweight process that runs in the background of your computer: a daemon.
Using yaq, a typical instrument has several daemons each supporting a particular component.
As illustrated in Figure \ref{fig:daemons}, each of these daemons has a single protocol which is typically composed of multiple traits as well as unique methods.

Paragraph to describe that methods return quickly and existence of busy.

In yaq, each component of an instrument can be developed and distributed separately.
For example, two very different instruments might happen to use the same temperature sensor.
Because the temperature sensor daemon is its own independent program, both instruments can benefit from the same daemon.
As yaq grows, the "ecosystem" of existing daemons means that future instruments become easier and easier to develop.

@@

\subsection{Clients} \label{sec:clients}

\begin{figure}
  \label{fig:acquisition}
  \includegraphics[width=\columnwidth]{./figures/acquisition.png}
  \caption{ A figure caption. The figure captions are automatically numbered.}
\end{figure}

yaq's RPC design invites the creation of multiple client libraries specialized for specific purposes.
Again, daemons can communicate simultaniously with multiple clients.
For a yaq instrument it is normal to have multiple clients running simultaniously, each focused on a particular purpose.
This modularity speeds development and increases overall flexibility of the system.

Figure \ref{fig:acquisition} illustrates using pseudocode what a very basic experimental acquisition might look like using yaq.
Here we have access to the protocols described in Figure \ref{fig:daemons}.
In the outermost for loop the variable ``pos'' is set to a series of different positions.
Each loop iteration involves setting a motor, launching a sensor measurement, then recording all independent and dependent variables.
Because all of these methods are trait-defined, this simple acquisition loop will run as-is for an arbitrary set of motors and sensors.
While this example is simplified, in practice we have found writing simple scripts to drive experiments to be a very productive approach.

@@

\section{Conclusion}

TODO

\begin{acknowledgments}
  TODO
\end{acknowledgments}

\section*{Data Availability Statement}

Data sharing is not applicable to this article as no new data were created or analyzed in this study.

\section*{References}

\nocite{*}
\bibliography{references}
